Transferability of clinical prediction models for early trauma care

### 2019-08-14 ManuscriptMarkdown.Rmd
Update ManuscriptMarkdown.Rmd

# Results
We analysed data from `r total.number.of.patients ` trauma patients to
investigate the effects of transfer on clinical prediction models (Table 1). The
total number of missing observations across all variables was `r total.number.of.na ` 
in the entire study cohort. The sample with the highest
percentage of missing observations was the `r highest.percent.missing.name `
sample with `r highest.percent.missing.number `% incomplete observations. The
variable with the highest number of missing values was `r name.variable.na `,
with `r number.variable.na ` missing values, or `r round(percent.variable.na)`% of 
the total values for this variable. The percentages of missing values for the other 
model predictors, was `r round(gcs.na$percentage)`% for GCS and `r round(sbp.na$percentage)`% 
for SBP.

**Table 1. Sample characteristics of multiple imputed data.** Data is presented as medians with interquartile range (IQR) or counts with % as applicable.
```{r, echo = FALSE, results = "asis"} 
## Table 1
options(knitr.kable.NA = '')
kable(high.low, format = "markdown")
kable(met.nonmet, format = "markdown")
kable(multi.single, format = "markdown")
# kable(complete.table.one, format = "markdown")
``` 
GCS, glasgow coma scale; SBP, systolic blood pressure; RR, respiratory rate; ISS, injury severity score; NISS, new injury severity score

## Development and validation
During model development, the number of imputations used for each data set was
`r high.low.imp`, `r metro.non.imp` and `r multi.single.imp` for the high and
low volume data set, the metropolitan and non-metropolitan data set, and the
multi- and single centre data set, respectively.

Table 2 shows model validation performance. The model with the lowest validation
mistriage rate was the `r colnames(lowest.val.mistriage)` model with a median
mistriage rate of `r lowest.val.mistriage`. Performance in terms of
undertriage and overtriage rate for the same model was `r val.undertriage`, and
`r val.overtriage` respectively. The worst validation performance (i.e. highest
mistriage rate) was found in the `r colnames(highest.val.mistriage)` sample, with
a model median mistriage rate of `r highest.val.mistriage`.

**Table 2. Model validation performance, i.e. performance when applied to the sample in which the model was created.** Data is presented as point estimates with 95% Uncertainty Intervals (95% UI).
```{r, echo = FALSE, results = "asis"} 
## Table 2
kable(table.2[,1:2], format = "markdown")
kable(table.2[,3:4], format = "markdown")
kable(table.2[,5:6], format = "markdown")
# kable(complete.table.two, format = "markdown")
```

## Comparison
Model performance after transfer was determined for each model being transferred
to the other sample in the data set, in all data sets (Table 3). When
transferred, the model with the lowest mistriage rate was the `r colnames(lowest.tran.mistriage)` 
model with a median mistriage rate of `r lowest.tran.mistriage`. The model with 
the highest mistriage rate after transfer was the `r colnames(highest.tran.mistriage)` 
with a median mistriage rate of `r highest.tran.mistriage`.

**Table 3. Model transfer performance, i.e. performance when transferred to the other sample in the same data set. ** Data is presented as point estimates with 95% Uncertainty Intervals (95% UI).  
```{r, echo = FALSE, results = "asis"} 
kable(table.3[,1:2], format = "markdown")
kable(table.3[,3:4], format = "markdown")
kable(table.3[,5:6], format = "markdown")
```

Table 4 shows the transferred model performance when compared to the validation
performance in the sample reciving the transferred model.
The highest mistriage rate when compared to the reciveing sample validation
model performance was found in the `r colnames(highest.comp.mistriage)` model,
with a median mistriage rate difference of `r highest.comp.mistriage`, meaning
this model performed worse when transferred to the other sample in the same data
set. In clinical terms, this model transfer means that among 100 trauma
patients, `r abs(worst.transfer$Transferred.mistriage.minus.buddy.local.mistriage.median) * 100` 
more patients would be wrongly classified as major or minor trauma.

**Table 4. Comparison mistriage, i.e. sample validation performance subtracted from transferred performance.** Data is presented as point estimates with 95% Uncertainty Intervals (95% UI).  
```{r, echo = FALSE, results = "asis"} 
kable(table.4[,1:2], format = "markdown")
kable(table.4[,3:4], format = "markdown")
kable(table.4[,5:6], format = "markdown")
```

# Discussion
This study aimed to assess how transfers of clinical prediction models for early
trauma care between different contexts within a single health system affect
mistriage rates. In all data sets, model transfer between samples resulted in
substantial uncertainty in its effect on model performance. When comparing
transferred performance to the receiving sample validation performance, the
resulting differences also showed considerable uncertainty.

The most notable effects on model performance following model transfer were seen
after transferring the `r colnames(highest.comp.mistriage)` model. This transfer
resulted in an increase of the mistriage rate by 
`r substr(format(highest.comp.mistriage), 1, 5)`.  Mainly contributing to this was
a marked increase in overtriage.  For this transfer, the 95% UI ranged from 
`r UI.top.bot[1]` to `r UI.top.bot[2]`, indicating a high level of uncertainty and
that our findings are compatible with a `r UI.top.bot[1]` decrease to a 
`r UI.top.bot[2]` increase in mistriage. The negative 95% UI of `r UI.top.bot[1]`
would signify an improvement in model performance. The higher 95% UI of 
`r UI.top.bot[2]` would mean a marked increase in model mistriage, compared to the
point estimate of `r substr(format(highest.comp.mistriage), 1, 5)`.

@@

Recognizing these risks and updating models accordingly could lead
to increased accuracy during trauma patient triage, potentially saving lives by
minimizing undertriage, and likely save resources by minimizing overtriage. With
registries such as SweTrau, analysis of model transfer could be performed
continuously to optimize models currently in use and advise which model
transfers might be detrimental (or potentially beneficial).

The transfer of clinical prediction models in trauma care has not previously
been studied extensively. In 2016, Gerdin et al. found model transfer to
adversely affect model performance in terms of calibration, but that this could
be improved by updating the model [@RN27]. This study did not examine model
performance in terms of calibration, neither were the effects of updating the
model explored.

External validation as the evaluation of a clinical prediction model in a
setting in which it was not originally developed has been studied more
extensively. Studies using simulated clinical prediction models based on
different predictors showed a decline in model performance when being externally
validated [@RN60; @RN61]. A study from 2018 examining the external validity of
prediction models for coronary artery disease (CAD) [@RN62], showed models
significantly underestimating the probability of CAD. The combined results of
this study, the studies on model transfer and the studies on external validation
would suggest caution when transferring models.

@@

This study used GCS, RR and SBP as predictors for 30-day
mortality. Additional predictors such as age, mechanism
of injury, or newer predictors like shock index might increase model
performance. Furthermore, we used 30-day mortality as outcome, recognizing
that late mortality, time in hospital, functional impairment, and morbidity are
also important outcome factors in the context of trauma care.

@@

## Conclusion
Model transfer resulted in large and varying levels of uncertainty in its effect
on mistriage. Therefore, model transfer can be unpredictable, potentially leading 
to increased mistriage, but in some cases also to a decrease in mistriage. 
Noticeably, transfer of the `r colnames(highest.comp.mistriage)` model lead to an 
increased mistriage of `r highest.comp.mistriage`. This observed effect was also 
associated with considerable uncertainty. Recognizing the limitations of this study, 
the authors believe further studies are warranted due to the potential economic 
costs and patient consequences of poor model transfers.

>>>>

# Results
We analysed data from `r total.number.of.patients ` trauma patients to
investigate the effects of transfer on clinical prediction models (Table 1). The
total number of missing observations across all variables was `r total.number.of.na ` 
in the entire study cohort. The sample with the highest
percentage of missing observations was the `r highest.percent.missing.name `
sample, with `r highest.percent.missing.number `% incomplete observations. The
variable with the highest number of missing values was the `r name.variable.na `,
with `r number.variable.na ` missing values, or `r round(percent.variable.na)`% of 
the total values for this variable. The percentages of missing values for the other 
model predictors were `r round(gcs.na$percentage)`% for GCS and `r round(sbp.na$percentage)`% 
for SBP.

**Table 1. Sample characteristics of multiple imputed data.** Data are presented as medians with interquartile range (IQR) or counts with %, as applicable.
```{r, echo = FALSE, results = "asis"} 
## Table 1
options(knitr.kable.NA = '')
kable(high.low, format = "markdown")
kable(met.nonmet, format = "markdown")
kable(multi.single, format = "markdown")
# kable(complete.table.one, format = "markdown")
``` 
GCS, Glasgow coma scale; SBP, systolic blood pressure; RR, respiratory rate; ISS, injury severity score; NISS, new injury severity score

## Development and validation
During model development, the number of imputations used for each data set was
`r high.low.imp`, `r metro.non.imp` and `r multi.single.imp` for the high- and
low-volume data set, the metropolitan and non-metropolitan data set, and the
multiple and single centre data set, respectively.

Table 2 shows the model validation performance. The model with the lowest validation
mistriage rate was the `r colnames(lowest.val.mistriage)` model, with a median
mistriage rate of `r lowest.val.mistriage`. The performance in terms of
undertriage and overtriage rate for the same model was `r val.undertriage` and
`r val.overtriage`, respectively. The worst validation performance (i.e. highest
mistriage rate) was found in the `r colnames(highest.val.mistriage)` sample, with
a model median mistriage rate of `r highest.val.mistriage`.

**Table 2. Model validation performance, i.e. performance when applied to the sample in which the model was created.** Data are presented as point estimates with 95% uncertainty intervals (95% UI).
```{r, echo = FALSE, results = "asis"} 
## Table 2
kable(table.2[,1:2], format = "markdown")
kable(table.2[,3:4], format = "markdown")
kable(table.2[,5:6], format = "markdown")
# kable(complete.table.two, format = "markdown")
```

## Comparison
Model performance after transfer was determined for each model after being transferred
to the other sample in the data set in all data sets (Table 3). When
transferred, the model with the lowest mistriage rate was the `r colnames(lowest.tran.mistriage)` 
model, with a median mistriage rate of `r lowest.tran.mistriage`. The model with 
the highest mistriage rate after transfer was the `r colnames(highest.tran.mistriage)` model 
with a median mistriage rate of `r highest.tran.mistriage`.

**Table 3. Model transfer performance, i.e. performance when transferred to the other sample in the same data set.** Data are presented as point estimates with 95% uncertainty intervals (95% UI).  
```{r, echo = FALSE, results = "asis"} 
kable(table.3[,1:2], format = "markdown")
kable(table.3[,3:4], format = "markdown")
kable(table.3[,5:6], format = "markdown")
```

Table 4 shows the transferred model performance when compared to the validation
performance in the sample receiving the transferred model.
The highest mistriage rate when compared to the receiving sample validation
model performance was found in the `r colnames(highest.comp.mistriage)` model,
with a median mistriage rate difference of `r highest.comp.mistriage`, meaning
this model performed worse when transferred to the other sample in the same data
set. In clinical terms, this model transfer means that among 100 trauma
patients, `r abs(worst.transfer$Transferred.mistriage.minus.buddy.local.mistriage.median) * 100` 
more patients would be wrongly classified as having major or minor trauma.

**Table 4. Comparison mistriage, i.e. sample validation performance subtracted from transferred performance.** Data are presented as point estimates with 95% uncertainty intervals (95% UI).  
```{r, echo = FALSE, results = "asis"} 
kable(table.4[,1:2], format = "markdown")
kable(table.4[,3:4], format = "markdown")
kable(table.4[,5:6], format = "markdown")
```

# Discussion
This study aimed to assess how transfers of clinical prediction models for early
trauma care between different contexts within a single health system affect
mistriage rates. In all data sets, model transfer between samples resulted in
substantial uncertainty in terms of its effect on model performance. When comparing the
transferred performance to the receiving sample validation performance, the
resulting differences also showed considerable uncertainty.

The most notable effects on model performance following model transfer were observed
after transferring the `r colnames(highest.comp.mistriage)` model. This transfer
resulted in an increase in the mistriage rate of 
`r substr(format(highest.comp.mistriage), 1, 5)`.  Mainly contributing to this was
a marked increase in overtriage.  For this transfer, the 95% UI ranged from 
`r UI.top.bot[1]` to `r UI.top.bot[2]`, indicating a high level of uncertainty, suggesting
that our findings are compatible with a `r UI.top.bot[1]` decrease to a 
`r UI.top.bot[2]` increase in mistriage. The negative 95% UI of `r UI.top.bot[1]`
would signify an improvement in model performance. The higher 95% UI of 
`r UI.top.bot[2]` would mean a marked increase in model mistriage compared to the
point estimate of `r substr(format(highest.comp.mistriage), 1, 5)`.

@@

Recognizing these risks and updating models accordingly could lead
to increased accuracy during trauma patient triage, potentially saving lives by
minimizing undertriage, and likely saving resources by minimizing overtriage. With
registries such as SweTrau, the analysis of model transfer could be performed
continuously to optimize the models currently in use and advise which model
transfers might be detrimental (or potentially beneficial).

The transfer of clinical prediction models in trauma care has not previously
been studied extensively. In 2016, Gerdin et al. found that model transfer
adversely affected model performance in terms of calibration but that this could
be improved by updating the model [@RN27]. This study did not examine model
performance in terms of calibration, nor were the effects of updating the
model explored.

External validation as the evaluation of a clinical prediction model in a
setting in which it was not originally developed has been studied more
extensively. Studies using simulated clinical prediction models based on
different predictors showed a decline in model performance when externally
validated [@RN60; @RN61]. A study from 2018 examining the external validity of
prediction models for coronary artery disease (CAD) [@RN62] showed that the models
significantly underestimated the probability of CAD. The combined results of
this study, the studies on model transfer and the studies on external validation
would suggest caution when transferring models.

@@

This study used GCS, RR and SBP as predictors for 30-day
mortality. Additional predictors such as age and mechanism
of injury or newer predictors such as the shock index might increase model
performance. Furthermore, we used 30-day mortality as an outcome, recognizing
that late mortality, time in hospital, functional impairment, and morbidity are
also important outcomes in the context of trauma care.

@@

## Conclusion
Model transfer resulted in large and varying levels of uncertainty in in terms of its effect
on mistriage. Therefore, model transfer can be unpredictable, potentially leading 
to increased mistriage but in some cases also to a decrease in mistriage. 
Noticeably, transfer of the `r colnames(highest.comp.mistriage)` model led to an 
increased mistriage of `r highest.comp.mistriage`. This observed effect was also 
associated with considerable uncertainty. Recognizing the limitations of this study, 
the authors believe that further studies are warranted due to the potential economic 
costs and patient consequences of poor model transfers.
